Considering scale in multilayer animal social networks

### 2020-06-28 paper/manuscript.Rmd
reflow, tidy

Social interactions exists across both temporal and spatial scales [@Whitehead_2008]. For
example, grooming requires close spatial proximity between conspecifics and
occurs over brief time periods [e.g. @Carter_2015], whereas social associations
over long time periods 
represent shared space use by members of the same social group [Figure
\@ref(fig:spacetime), @Franks_2009]. Individuals must share space to 
interact or associate and thus have overlapping
home ranges [@Vander_2014]. Home
range overlap is an example of how animals share space over coarser spatial and
temporal scales <!-- TODO to create their social environment -->[e.g. @Piza_Roca_2018]. Although the social environment
clearly scales spatially and temporally, it remains unclear whether coarser
scales of social interactions, such as social association and home range overlap, vary
predictably [@Castles_2014; @Farine_2015a]. 
We therefore define social scale as the explicit definition 
and measurement of social relationships [@Farine_2015a].

@@

### Network layer construction
We generated proximity-based social networks using the R package `spatsoc`
[@Robitaille_2019] in `r substr(version$version.string, 1, 15)` [@R_Core_Team]. We assumed
individuals were associating if simultaneous GPS fixes were within the spatial
distance threshold of one another. Typically for ungulates and other gregarious
mammals, the ‘chain rule’ is used for group assignment [@Croft_2008]. For GPS
data, the chain rule is applied by assigning a group identifier to the union of
buffered GPS fixes at each time step. As such, individuals in a group are within
the social distance threshold of at least one other though not necessarily all
other individuals [@Robitaille_2019]. <!-- TODO: check jwt comment--> Group assignment based on the chain rule
has previously been applied to caribou using a temporal threshold of 5 minutes
and a social distance threshold of 50 m [@Peignier_2019; @Lesmerises_2018]. In
all networks, individual caribou were represented as nodes and associations between
individuals in a given network layer were represented as intralayer
edges.

@@

The temporal multilayer network indicated within-year time windows where social 
network metrics increased. Edge overlap was higher in time windows 1, 6-11 and 
15-20 than in time windows 2-5 and 12-15 
(Figure \@ref(fig:temp) A). Time windows of higher edge overlap, between 
approximately October and June, also had higher graph strength 
compared to time windows of lower edge overlap, between approximately June 
and October (Figure \@ref(fig:temp) C). Graph strength peaked across the time 
series between January 2018 and April 2018. 
These results suggest that ordinal multilayer networks can effectively 
capture discrete time windows of aggregation and dissaggregation related 
to resource availability, and highlight the importance of carefully 
selecting the temporal scale of analysis. <!-- TODO see comment-->

As the number of observations used to generate multilayer networks increased,
individuals became more connected and variance in metrics decreased.
Edge overlap increased across all three habitat
layers with increasing number of observations (Figure \@ref(fig:nobs) B).
Similarly, multidegree for all individuals increased quickly between 10 and 100
observations, and continued to increase until 1000 observations (Figure
\@ref(fig:nobs) C). Graph strength across individuals in all three habitat layers was highly
variable at low number of observations and variance decreased after ~100 observations
(Figure \@ref(fig:nobs) A). <!--TODO double check-->Since the variance in these 
multilayer metrics decreases after a certain number
of observations, this sensitivity method could be useful for determining the
number of observations necessary for sufficiently describing social dynamics
across contexts.

@@

# Discussion
<!--
1. intro
TODO: somewhere ref:ml
-->

<!-- intro -->
Animal social systems are characterized and influenced by scale. Three types of
scale to consider are social, spatial, and temporal scale. Social scale defines
the type and measurement of social relationships and directly impacts observed
social connectivity between individuals. Spatial scale can reflect the
resolution of the habitat within which social interactions occur. Finally,
temporal scale is related to both study design and seasonal differences in
social processes. By partitioning social associations into discrete contexts
using multilayer networks, we highlight the influence of social, spatial and
temporal scales on animal social systems, demonstrating the importance of
considering biologically relevant and robust scales.
<!-- intro -->

<!-- social scale -->
Social scale is an essential consideration for social network analysis
[@Castles_2014; @Carter_2015; @Farine_2015a]. For example, networks can be
constructed based on fine-scale social interactions (e.g. grooming or
aggression), social associations (e.g. group membership or proximity), or
spatial or home range overlap. Here, we define social associations using a
series of social distance thresholds to generate proximity-based social
networks. Across network layers, increasing the social distance threshold
resulted in an increase in graph strength (Figure \@ref(fig:socres)).
Specifically, the strength of social associations across habitat layers, with
open and foraging habitats demonstrating a more pronounced increase in graph
strength as well as greater variation across individuals. These results
emphasize that multilayer network metrics are directly influenced by social
scale. An extension of this analysis could be to incorporate behavioural states
to explore the influence of social scale on patterns of association or
interaction across habitat and behavioural layers [e.g. using focal observations
or hidden Markov Models see @Muller_2018; @Jones_2020].
<!-- social scale -->

<!-- spatial -->
Habitat selection, the use of available resources or landscape features by
animals [@Manly_1993; @Boyce_1999], is a scale-dependent process [@Mayor_2009].
Social structure and habitat selection are intimately related and individuals
must share space to interact [@Webber_2018]. We found edge overlap was stable in
open and forest layers but decreasing in foraging layers with increasing spatial
scale (Figure \@ref(fig:lcres) D). Graph strength had greater individual
variation in open and foraging layers than in forest layers (Figure
\@ref(fig:lcres) E). We observed a decrease in the proportion of relocations in
foraging habitat with increasing spatial resolution (`r count[res == 30 & lc == 3, percent]`% at 30 m - `r count[res == 1000 & lc == 3, percent]`% at 1000
m). As the foraging layer had similar edge overlap and graph strength as the
open layer at 30 m spatial resolution despite having lower abundance on the
landscape (foraging: 12.8%, open: 36.9%), this highlights the importance
of considering spatial scale when including habitat in multilayer networks.
Johnson [-@Johnson_1980] identified four orders of habitat selection: (1) the
geographical range of a species, (2) the home range of an individual, the (3)
habitat patches within the home range and (4) specific resources within a
habitat patch. Kie et al. [-@Kie_2002] found that, independently, third order of
selection was insufficient for capturing landscape attributes selected for by
mule deer (*Odocoileus hemionus*), at a scale larger than their home range size.
Scales of habitat selection could be relevant for multilayer networks and our
integration of habitat within a multilayer context could be an important first
step (see Figure \@ref(fig:lcres)). Specifically, if individuals select habitats
at the home range scale, it may be relevant to assess social structure using
home range overlap. By contrast, in cases where individuals select habitats at
the patch-scale, fine-scale measures of sociality may be more relevant. Moving
past this hierarchical perspective, Mayor et al. [-@Mayor_2007] emphasized the
importance of considering a broad continuum of scales to reveal scale-dependent
selection and avoid the bias of predefined scales assumed to be associated with
certain behaviours.
<!-- spatial -->

<!-- temp -->
The emergent relationship between habitat and sociality has been explored by the
temporal distribution and phenology of resources and use of social information.
We found increased edge overlap and graph strength in fall, winter and spring
(approximately October - June) than in summer (approximately June - October)
(Figure \@ref(fig:temp) A, C). These periods of increased sociality correspond
to seasons of heterogeneous resource availability [Peigner_2019]. This analysis
could be extended using time-aggregated networks and a moving-window approach to
time window size and sampling effort [@Farine_2017; @Bonnell_2020]. By
explicitly considering the temporal context of social interactions in separate
layers, multilayer networks provide a detailed and dynamic perspective of animal
social networks. For example, female mandrill (*Mandrillus sphinx*) sociality is
more completely represented by temporal multilayer networks than aggregated
networks [@Pereira_2020]. Evans et al. [-@Evans_2020] observed seasonal
differences in the social structure of mice (*Mus musculus domesticus*) related
to temperature and breeding activity. Finally, it is crucial to include
sufficient data within sampling periods or time windows to construct robust
networks [@Farine_2017; Proskurnikov_2017]. With increasing number of
observations, we found variance decreased in edge overlap and multidegree across
habitat layers as well as in graph strength within habitat layers (Figure
\@ref(fig:nobs)). As more observations were added, rarer relationships were
recorded demonstrating the value of this form of sensitivity analysis.
<!-- temp -->

<!-- Finally, due to the autocorrelated nature of biologging data it is -->
<!-- essential to ensure social association data area non-randomly distributed -->
<!-- [@Farine_2017b]. -->

<!-- conclusion -->
Recognition of scale as a fundamental concept in ecology [@Levin_1992] has lead
to an increased attention to the scale at which ecological processes are
observed [@Schneider_2001; @Chave_2013]. Recent technological advances in
remotely sensed imagery [@Drusch_2012] and animal tracking devices
[@Borger_2020; @B_rger_2016; @Wikelski_2007] have increased the availability of
complex, high resolution data. We highlight ways in which scale should be
considered when building social networks from both observed and remotely-sensed
data. Our multilayer network analysis of caribou sociality incorporating high
resolution GPS telemetry data and spatial land cover data is grounded in the
context of observed animal behaviour as the scales we chose to explore, e.g.,
number of observations and sociality across space and time, are ones that have
also traditionally been done in ethology.

@@

We show that multilayer social networks are (1) scale-dependent, and (2) useful
for identifying spatially or temporally specific social associations. When
employing multilayer networks, it is critical to consider a broad continuum of
social, spatial and temporal scales to fully capture ecological processes and to
establish a clear link between biologically relevant scales of observation.
<!-- conclusion -->

@@

# Box 1
Social phenotype has been increasingly linked to individual fitness, including
in primates [@Brent_2017; @Thompson_2019], dolphins [@Stanton_2012;
@Stanton_2011], and birds [@McDonald_2007; @Royle_2012]. Social and temporal
scales have emerged as important variables for understanding the fitness
consequences of social network position [@Almeling_2016; @Berger_2015;
@Brent_2017; @Holekamp_2011].

>>>>

Social interactions exists across both temporal and spatial scales
[@Whitehead_2008]. For example, grooming requires close spatial proximity
between conspecifics and occurs over brief time periods [e.g. @Carter_2015],
whereas social associations over long time periods represent shared space use by
members of the same social group [Figure \@ref(fig:spacetime), @Franks_2009].
Individuals must share space to interact or associate and thus have overlapping
home ranges [@Vander_2014]. Home range overlap is an example of how animals
share space over coarser spatial and temporal scales [e.g. @Piza_Roca_2018].
Although the social environment clearly scales spatially and temporally, it
remains unclear whether coarser scales of social interactions, such as social
association and home range overlap, vary predictably [@Castles_2014;
@Farine_2015a]. We therefore define social scale as the explicit definition and
measurement of social relationships [@Farine_2015a].

@@

### Network layer construction
We generated proximity-based social networks using the R package `spatsoc`
[@Robitaille_2019] in `r substr(version$version.string, 1, 15)` [@R_Core_Team]. We assumed
individuals were associating if simultaneous GPS fixes were within the spatial
distance threshold of one another. Typically for ungulates and other gregarious
mammals, the ‘chain rule’ is used for group assignment [@Croft_2008]. For GPS
data, the chain rule is applied by assigning a group identifier to the union of
buffered GPS fixes at each time step. As such, individuals in a group are within
the social distance threshold of at least one other though not necessarily all
other individuals [@Robitaille_2019]. Group assignment based on the chain rule
has previously been applied to caribou using a temporal threshold of 5 minutes
and a social distance threshold of 50 m [@Peignier_2019; @Lesmerises_2018]. In
all networks, individual caribou were represented as nodes and associations between
individuals in a given network layer were represented as intralayer
edges.

@@

The temporal multilayer network indicated within-year time windows where social
network metrics increased. Edge overlap was higher in time windows 1, 6-11 and
15-20 than in time windows 2-5 and 12-15 (Figure \@ref(fig:temp) A). Time
windows of higher edge overlap, between approximately October and June, also had
higher graph strength compared to time windows of lower edge overlap, between
approximately June and October (Figure \@ref(fig:temp) C). Graph strength peaked
across the time series between January 2018 and April 2018. These results
highlight the importance of carefully selecting the temporal scale of analysis
to effectively capture within- and across-year variation in sociality.

As the number of observations used to generate multilayer networks increased,
individuals became more connected and variance in metrics decreased. Edge
overlap increased across all three habitat layers with increasing number of
observations (Figure \@ref(fig:nobs) B). Similarly, multidegree for all
individuals increased quickly between 10 and 100 observations, and continued to
increase until 1000 observations (Figure \@ref(fig:nobs) C). Graph strength
across individuals in all three habitat layers was highly variable at low number
of observations and variance decreased after ~100 observations (Figure
\@ref(fig:nobs) A). Since the variance in these multilayer metrics decreases
after a certain number of observations, this sensitivity method could be useful
for determining the number of observations necessary for sufficiently describing
social dynamics across contexts.

@@

# Discussion
Animal social systems are characterized and influenced by scale. Three types of
scale to consider are social, spatial, and temporal scale. Social scale defines
the type and measurement of social relationships and directly impacts observed
social connectivity between individuals. Spatial scale can reflect the
resolution of the habitat within which social interactions occur. Finally,
temporal scale is related to both study design and seasonal differences in
social processes. By partitioning social associations into discrete contexts
using multilayer networks, we highlight the influence of social, spatial and
temporal scales on animal social systems, demonstrating the importance of
considering biologically relevant and robust scales.

Social scale is an essential consideration for social network analysis
[@Castles_2014; @Carter_2015; @Farine_2015a]. For example, networks can be
constructed based on fine-scale social interactions (e.g. grooming or
aggression), social associations (e.g. group membership or proximity), or
spatial or home range overlap. Here, we define social associations using a
series of social distance thresholds to generate proximity-based social
networks. Across network layers, increasing the social distance threshold
resulted in an increase in graph strength (Figure \@ref(fig:socres)).
Specifically, the strength of social associations across habitat layers, with
open and foraging habitats demonstrating a more pronounced increase in graph
strength as well as greater variation across individuals. These results
emphasize that multilayer network metrics are directly influenced by social
scale. An extension of this analysis could be to incorporate behavioural states
to explore the influence of social scale on patterns of association or
interaction across habitat and behavioural layers [e.g. using focal observations
or hidden Markov Models see @Muller_2018; @Jones_2020].

Habitat selection, the use of available resources or landscape features by
animals [@Manly_1993; @Boyce_1999], is a scale-dependent process [@Mayor_2009].
Social structure and habitat selection are intimately related and individuals
must share space to interact [@Webber_2018]. We found edge overlap was stable in
open and forest layers but decreasing in foraging layers with increasing spatial
scale (Figure \@ref(fig:lcres) D). Graph strength had greater individual
variation in open and foraging layers than in forest layers (Figure
\@ref(fig:lcres) E). We observed a decrease in the proportion of relocations in
foraging habitat with increasing spatial resolution (`r count[res == 30 & lc == 3, percent]`% at 30 m - `r count[res == 1000 & lc == 3, percent]`% at 1000
m). As the foraging layer had similar edge overlap and graph strength as the
open layer at 30 m spatial resolution despite having lower abundance on the
landscape (foraging: 12.8%, open: 36.9%), this highlights the importance
of considering spatial scale when including habitat in multilayer networks.
Johnson [-@Johnson_1980] identified four orders of habitat selection: (1) the
geographical range of a species, (2) the home range of an individual, the (3)
habitat patches within the home range and (4) specific resources within a
habitat patch. Kie et al. [-@Kie_2002] found that, independently, third order of
selection was insufficient for capturing landscape attributes selected for by
mule deer (*Odocoileus hemionus*), at a scale larger than their home range size.
Scales of habitat selection could be relevant for multilayer networks and our
integration of habitat within a multilayer context could be an important first
step (see Figure \@ref(fig:lcres)). Specifically, if individuals select habitats
at the home range scale, it may be relevant to assess social structure using
home range overlap. By contrast, in cases where individuals select habitats at
the patch-scale, fine-scale measures of sociality may be more relevant. Moving
past this hierarchical perspective, Mayor et al. [-@Mayor_2007] emphasized the
importance of considering a broad continuum of scales to reveal scale-dependent
selection and avoid the bias of predefined scales assumed to be associated with
certain behaviours.

The emergent relationship between habitat and sociality has been explored by the
temporal distribution and phenology of resources and use of social information.
We found increased edge overlap and graph strength in fall, winter and spring
(approximately October - June) than in summer (approximately June - October)
(Figure \@ref(fig:temp) A, C). These periods of increased sociality correspond
to seasons of heterogeneous resource availability [Peigner_2019]. This analysis
could be extended using time-aggregated networks and a moving-window approach to
time window size and sampling effort [@Farine_2017; @Bonnell_2020]. By
explicitly considering the temporal context of social interactions in separate
layers, multilayer networks provide a detailed and dynamic perspective of animal
social networks. For example, female mandrill (*Mandrillus sphinx*) sociality is
more completely represented by temporal multilayer networks than aggregated
networks [@Pereira_2020]. Evans et al. [-@Evans_2020] observed seasonal
differences in the social structure of mice (*Mus musculus domesticus*) related
to temperature and breeding activity. Finally, it is crucial to include
sufficient data within sampling periods or time windows to construct robust
networks [@Farine_2017; Proskurnikov_2017]. With increasing number of
observations, we found variance decreased in edge overlap and multidegree across
habitat layers as well as in graph strength within habitat layers (Figure
\@ref(fig:nobs)). As more observations were added, rarer relationships were
recorded demonstrating the value of this form of sensitivity analysis.

Recognition of scale as a fundamental concept in ecology [@Levin_1992] has lead
to an increased attention to the scale at which ecological processes are
observed [@Schneider_2001; @Chave_2013]. Recent technological advances in
remotely sensed imagery [@Drusch_2012] and animal tracking devices
[@Borger_2020; @B_rger_2016; @Wikelski_2007] have increased the availability of
complex, high resolution data. We highlight ways in which scale should be
considered when building social networks from both observed and remotely-sensed
data. Our multilayer network analysis of caribou sociality incorporating high
resolution GPS telemetry data and spatial land cover data is grounded in the
context of observed animal behaviour as the scales we chose to explore, e.g.,
number of observations and sociality across space and time, are ones that have
also traditionally been done in ethology.

@@

We show that multilayer social networks are (1) scale-dependent, and (2) useful
for identifying spatially or temporally specific social associations. When
employing multilayer networks, it is critical to consider a broad continuum of
social, spatial and temporal scales to fully capture ecological processes and to
establish a clear link between biologically relevant scales of observation.

# Box 1
<!-- Finally, due to the autocorrelated nature of biologging data it is -->
<!-- essential to ensure social association data area non-randomly distributed -->
<!-- [@Farine_2017b]. -->

@@

Social phenotype has been increasingly linked to individual fitness, including
in primates [@Brent_2017; @Thompson_2019], dolphins [@Stanton_2012;
@Stanton_2011], and birds [@McDonald_2007; @Royle_2012]. Social and temporal
scales have emerged as important variables for understanding the fitness
consequences of social network position [@Almeling_2016; @Berger_2015;
@Brent_2017; @Holekamp_2011].
